---
title: "Crop Water Stress Prediction"
format: html
---

```{r}
#| output: FALSE
library(vroom)
library(tidyverse)
library(spmodel)   # splm (spatial linear models)
library(viridis)   # colorblind friendly colors
library(patchwork) # put graphs together
library(FRK)       # (Fixed Rank Kriging) --> Spatial interpolation & modeling (big data)
library(caret)     # trainControl for Cross-Validation
```

Read in the data:
```{r}
crop_na <- vroom("PotatoCWSI.csv", show_col_types = FALSE)

crop <- crop_na |>
  filter(!is.na(CWSI))

head(crop)
```

```{r}
croplonglat <- crop |> select(POINT_X, POINT_Y) 
observed_loc_matrix <- as.matrix(croplonglat)

## Number of Spatial Features
K <- 20

## "Centers" of Features
centers <- fields::cover.design(observed_loc_matrix, nd=K)$design 

## Use 1.5 rule of thumb for scale
dist_between_centers <- fields::rdist(centers) # KxK matrix for distances
the_scale <- 1.5*max(apply(dist_between_centers,1,function(x){sort(x)[2]})) # tau

## Define spatial features
spatial_features <- local_basis(
  manifold=plane(),
  loc=centers,
  scale=rep(the_scale, K),
  type="bisquare" #or "Gaussian" or "exp"
  ) %>%
  eval_basis(observed_loc_matrix) %>%
  as.matrix()
colnames(spatial_features) <- paste0("SF", 1:K) # name spatial features

## Bind new spatial features in data frame
spatial_df <- bind_cols(crop, spatial_features)
head(spatial_df)
```

# make our linear model

```{r}
sf_names <- paste0("SF", 1:20)
cwsi_all_SF <- as.formula(paste("CWSI ~ SLOPE + TWI + ASPECT + ECA_SHALLOW + NDVI + ", 
                             paste(sf_names, collapse = " + ")))

spatial_lm <- splm(formula = cwsi_all_SF,
                   data = spatial_df,
                   spcov_type = "exponential",
                   xcoord = POINT_X, 
                   ycoord = POINT_Y) 
```

# compare to reduced models
without spatial features
```{r}
reduced_lm <- splm(formula = CWSI ~ SLOPE + TWI + ASPECT + ECA_SHALLOW + NDVI,
                   data = spatial_df,
                   spcov_type = "exponential",
                   xcoord = POINT_X, 
                   ycoord = POINT_Y) 
lmtest::lrtest(reduced_lm, spatial_lm)
```
without ASPECT + ECA_SHALLOW + NDVI
```{r}
cwsi_slope_twi_SF <- as.formula(paste("CWSI ~ SLOPE + TWI + ", 
                             paste(sf_names, collapse = " + ")))

reduced_lm2 <- splm(formula = cwsi_slope_twi_SF,
                   data = spatial_df,
                   spcov_type = "exponential",
                   xcoord = POINT_X, 
                   ycoord = POINT_Y) 
lmtest::lrtest(reduced_lm2, spatial_lm)
```

```{r}
reduced_lm3 <- splm(formula = CWSI ~ 1,
                   data = spatial_df,
                   spcov_type = "exponential",
                   xcoord = POINT_X, 
                   ycoord = POINT_Y) 

just_SF <- as.formula(paste("CWSI ~ ", 
                             paste(sf_names, collapse = " + ")))
reduced_lm4 <- splm(formula = just_SF,
                   data = spatial_df,
                   spcov_type = "exponential",
                   xcoord = POINT_X, 
                   ycoord = POINT_Y) 

lmtest::lrtest(reduced_lm3, reduced_lm4)
```
low p-val, but clear spatial structure

# Line Assumptions
```{r}
decorr_resid <- rstandard(spatial_lm)
fitted_vals <- fitted(spatial_lm, what = "mu")
```

```{r}
# library(nlme)
# # Fit a model that accounts for the spatial trend (e.g., Exponential)
# spatial_gls <- gls(cwsi_all_SF, 
#                    correlation = corExp(form = ~ POINT_X + POINT_Y, nugget = TRUE), 
#                    data = spatial_df)
# 
# # To see if it worked, extract "normalized" residuals
# # These are pre-multiplied by the inverse of the correlation matrix
# decorr_resid <- residuals(spatial_gls, type = "normalized")
# fitted_vals <- fitted(spatial_gls, what = "mu")
```

Independence
```{r}
sp1 <- spatial_df |>
  mutate(resids = decorr_resid)
sp1 |>
  esv(formula=resids~1, xcoord=POINT_X, ycoord=POINT_Y) |>
  plot()
```
Normality
```{r}
ks <- ks.test(decorr_resid, "pnorm", mean=0, sd=1)
pval <- formatC(ks$p.value, digits = 4)

hr <- ggplot(data.frame(resids = decorr_resid), aes(x = resids)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "orchid",
                 alpha = 0.7, color = "orchid") +
  stat_function(fun = dnorm, args = list(mean = mean(decorr_resid), 
                                         sd = sd(decorr_resid)),
                linewidth = 0.7, color = "darkred", linetype = "dashed") +
  labs(title = paste0("Histogram of Decorrelated Residuals", 
                      "\nKS test p-value: ", pval),
       x = "Residuals",
       y = "Density") +
  theme_minimal()
hr
```
Equal Variance
```{r}
bp_test <- lmtest::bptest(
  decorr_resid ~ CWSI,
  data = spatial_df
)
bp_pval <- formatC(bp_test$p.value, digits = 4)

fr <- ggplot(data.frame(fitted = fitted_vals, resid = decorr_resid),
       aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Fitted Values",
       y = "Residuals",
       title = paste0("Fitted vs Decorrelated Residuals\n",
                    "BP Test p-value: ", bp_pval)) +
  theme_minimal()
fr
```


# Cross Validation <-- for description section
Measure of how well your model predicts (RMSE of like 0.3 or something)
10 fold Cross Validation:
```{r}
n <- nrow(spatial_df)
k <- 10

folds <- sample(rep(1:k, length.out = n))

rmse_vals <- numeric(k)

for (i in 1:k) {
  
  # split data
  train_data <- spatial_df[folds != i, ]
  test_data  <- spatial_df[folds == i, ]
  
  # fit model on training
  model_cv <- splm(
    formula = cwsi_all_SF,
    data = train_data,
    spcov_type = "exponential",
    xcoord = POINT_X,
    ycoord = POINT_Y
  )
  
  # predict on test
  preds <- predict(model_cv, newdata = test_data)
  
  # compute RMSE
  rmse_vals[i] <- sqrt(mean((test_data$CWSI - preds)^2))
}

# median RMSE
median_rmse <- median(rmse_vals)
median_rmse
```





# Predictions

```{r}
pred_grid <- crop_na %>% filter(is.na(CWSI))
# Prepare the spatial features for the NA locations
pred_basis <- local_basis(
  manifold = plane(), 
  loc = centers, 
  scale = rep(the_scale, K), 
  type = "bisquare") %>%
  eval_basis(as.matrix(pred_grid[, c("POINT_X", "POINT_Y")])) %>%
  as.matrix()

colnames(pred_basis) <- paste0("SF", 1:K)

# Combine the given info and the new spatial features
pred_data_final <- bind_cols(pred_grid, as.data.frame(pred_basis))

# Predict with Intervals --> returns a matrix with columns: fit, lwr, upr
preds <- predict(spatial_lm, newdata = pred_data_final, interval = "prediction", level = 0.95)

# Attach back to our grid
pred_grid_results <- pred_data_final %>%
  mutate(
    fit = preds[,1],
    lwr = preds[,2],
    upr = preds[,3]
  )
```

```{r}
save(spatial_df, spatial_lm, pred_grid_results, file = "./potato.RData")
```

```{r}
# shared scale
min_val <- min(c(pred_grid_results$fit, pred_grid_results$lwr, pred_grid_results$upr), na.rm = TRUE)
max_val <- max(c(pred_grid_results$fit, pred_grid_results$lwr, pred_grid_results$upr), na.rm = TRUE)
shared_limits <- c(min_val, max_val)

fitplot <- ggplot(pred_grid_results, aes(x = POINT_X, y = POINT_Y)) +
  geom_point(aes(color = fit), size = 1) +
  scale_color_viridis_c(limits = shared_limits) + # Syncing here
  labs(title = "Predicted CWSI (Fit)", x = "X coord", y = "Y coord", color = "CWSI") +
  theme_minimal()

lwrplot <- ggplot(pred_grid_results, aes(x = POINT_X, y = POINT_Y)) +
  geom_point(aes(color = lwr), size = 1) +
  scale_color_viridis_c(limits = shared_limits) + # Syncing here
  labs(title = "Lower 95% Bound", x = "X coord", y = "Y coord", color = "CWSI") +
  theme_minimal()

uprplot <- ggplot(pred_grid_results, aes(x = POINT_X, y = POINT_Y)) +
  geom_point(aes(color = upr), size = 1) +
  scale_color_viridis_c(limits = shared_limits) + # Syncing here
  labs(title = "Upper 95% Bound", x = "X coord", y = "Y coord", color = "CWSI") +
  theme_minimal()

(fitplot / (lwrplot + uprplot)) + 
  plot_layout(heights = c(4, 2), guides = "collect")& 
  coord_fixed() & 
  theme(legend.position = "right")
```



